---
title: "Diagnosing your R setup"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{diagnosing-your-r-setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

If you are struggling with R, Git or `renv` on your DfE laptop, the `diagnostic_test()` function in `dfeR` can help you work out what is going on. It runs a series of checks against your machine and reports which settings look healthy and which are likely to be causing you trouble.

By default the function is read-only and only tells you what it has found. If you would like it to attempt to fix things for you, you can run it with `clean = TRUE`, but you should always run it once without that argument first so that you can see what would change before anything is altered.

```{r run_diagnostic, eval = FALSE}
dfeR::diagnostic_test()
```

## Reading the summary

Each check finishes with a status. The summary line at the bottom of the output tells you how many checks fell into each category:

- **PASS** means the check found nothing to worry about.
- **FAIL** means the check found a setting that is likely to cause problems. Running with `clean = TRUE` will attempt to fix most of these.
- **FIXED** means the check found a problem and was able to fix it during this run (you only see this when you call the function with `clean = TRUE`).
- **INFO** is for checks that simply report information back to you, such as the path to your global Git config file. These are never failures.

If everything has come back as **PASS** or **INFO**, you can stop here. If anything has failed, the rest of this vignette walks through each check in turn and explains what to do.

## Proxy settings

`check_proxy_settings()` looks for proxy entries in your global Git config (`http.proxy`, `https.proxy`) and in the system environment variables (`http_proxy`, `https_proxy`, `no_proxy`).

These settings used to be needed inside the DfE network years ago, but they now prevent Git from connecting to GitHub and Azure DevOps. If you see an error like:

```
fatal: unable to access 'https://...': Could not resolve proxy: mwg.proxy.ad.hq.dept
```

then this is almost certainly what is causing it.

Running `check_proxy_settings(clean = TRUE)` will clear the Git config entries and unset the environment variables for the current R session. If your proxy environment variables were set permanently in Windows (using `setx` in the past), they will come back when you open a new R session. To remove them permanently, open a Windows terminal and run:

```
setx http_proxy ""
setx https_proxy ""
```

`setx` only affects *new* shell and RStudio sessions, so close and reopen RStudio after running this for the change to be picked up.

The equivalent command for the Git config, which you can run from your R console or a terminal, is:

```
git config --global --unset http.proxy
git config --global --unset https.proxy
```

## Git SSL verification

`check_git_sslverify()` looks at the `http.sslverify` and `https.sslverify` keys in your global Git config. A historic DfE setup script set these to `false` to work around SSL certificate problems on the internal network. That is no longer necessary, and leaving SSL verification disabled is not something you want on a long-term basis.

Running `check_git_sslverify(clean = TRUE)` will set both keys back to `true`.

If you turned SSL verification off recently because you were seeing a genuine SSL certificate error from R, Git or another tool, turning it back on will bring that error back. The right next step in that case is to contact the [Statistics Development Team](mailto:statistics.development@education.gov.uk) with the command you were running and the error message you saw, so we can work with the Network Operations team to unblock the underlying request.

## Global .gitconfig location

`check_gitconfig_location()` reports the path that Git is using for your global config file. It does not change anything; it only tells you where Git, R and RStudio think your global config lives.

This check fails when your global `.gitconfig` is inside OneDrive. When that happens, Git, R and RStudio can disagree about which config file is in effect, and changes you make in one tool may not be picked up by the others.

There is no automatic fix for this one because it is a Windows-level change rather than something R can do for you. The recommended location is `C:/Users/<username>/.gitconfig`. You will need to move the file there yourself, and you may also need to set the `HOME` environment variable on your machine so that everything agrees on where your home directory is.

## GITHUB_PAT

`check_github_pat()` looks at the `GITHUB_PAT` system environment variable. When this variable is set, R can fail to install packages from GitHub (including `dfeR` itself, and `dfeshiny`) with an error like:

```
ERROR [curl: (22) The requested URL returned error: 401]
```

For your safety, the diagnostic only ever shows the last four characters of the value, and the returned object only contains the masked version. The raw token never appears in the output, so it is safe to share the result of `diagnostic_test()` with a colleague.

Running `check_github_pat(clean = TRUE)` clears the variable for the current R session. If the variable comes back the next time you start R, something on your machine is setting it for you, often a credential manager or another bit of software, and you will need to find and disable whichever that is. The function prints a reminder about this when it clears the variable.

## renv download method (.Renviron)

`check_renv_download_method()` looks for the `RENV_DOWNLOAD_METHOD` entry in your `~/.Renviron` file. `renv` can download packages using either `curl` or `wininet`, and `wininet` does not work from within the DfE network. If you see a `renv` error like "wininet" failing to download packages, this is the setting to change.

Running `check_renv_download_method(clean = TRUE)` will rewrite `~/.Renviron` to set `RENV_DOWNLOAD_METHOD="curl"`, removing any previous assignments of the same variable. If you would rather change the file yourself, you can open it in RStudio with:

```{r edit_renviron, eval = FALSE}
usethis::edit_r_environ()
```

and add the line:

```
RENV_DOWNLOAD_METHOD="curl"
```

You will need to restart R after editing `.Renviron` for the change to take effect.

## RENV_DOWNLOAD_FILE_METHOD

`check_renv_dl_file_method()` looks at the `RENV_DOWNLOAD_FILE_METHOD` system environment variable. The legacy DfE setup script set this to `wininet`, which breaks `renv` outside the DfE network and is no longer needed.

Running `check_renv_dl_file_method(clean = TRUE)` will unset the variable for the current R session. If the variable was set permanently with `setx`, this clears it for the session you are in but it will come back the next time you start R. To remove it permanently, open a Windows terminal and run:

```
setx RENV_DOWNLOAD_FILE_METHOD ""
```

As with the proxy variables, `setx` only takes effect in new shell and RStudio sessions, so close and reopen RStudio afterwards.

## RTools

`check_rtools()` looks for `make` on your `PATH`, which on Windows tells us whether RTools is installed. Some R packages need to be compiled from source, and on Windows that requires RTools. A missing `make` will often show up as a "cannot find make" error during `renv::restore()` for older package versions that have to be built from source.

This check is reported as `INFO` rather than as a failure when `make` is missing, because plenty of users do not need to compile packages from source. If you do need it, the easiest way to get RTools at the DfE is from the Software Centre. Pick the version that matches your version of R, install it, then close and reopen RStudio.

If you cannot find RTools in the Software Centre, you can also install it directly from the [CRAN RTools page](https://cran.r-project.org/bin/windows/Rtools/).

## .Renviron / .Rprofile location

`check_renv_rprof_location()` reports where R is resolving `~/.Renviron` and `~/.Rprofile`, and the values of `HOME`, `R_USER` and `path.expand("~")`. It does not change anything; like the gitconfig check, it tells you where things are.

The check fails when your home directory has been redirected into OneDrive. When that happens, R, RStudio and Git may resolve `~` inconsistently, and `.Renviron` or `.Rprofile` changes you make in one tool may not be picked up by another. As with the gitconfig check, the fix is a Windows-level change rather than something R can do automatically. You will usually need to set the `HOME` environment variable explicitly to a path outside OneDrive (for example `C:/Users/<username>`) and make sure your `.Renviron` and `.Rprofile` live in that location.

If your home directory is not in OneDrive, this check will report `INFO` and you do not need to do anything.

## Sharing the full output with support

If you have worked through the failed checks and are still stuck, you can run the diagnostic in `full` mode, which appends a session dump containing `renv::diagnostics()`, `Sys.getenv()` and `options()`:

```{r run_diagnostic_full, eval = FALSE}
dfeR::diagnostic_test(full = TRUE)
```

The function masks values for any environment variable whose name contains `TOKEN`, `SECRET`, `KEY`, `PAT`, `PASSWORD` or `CRED`, so your `GITHUB_PAT` and similar tokens are not shown in full.

### Troubleshooting

A few common issues come up alongside the ones the diagnostic checks for. These are not currently part of `diagnostic_test()`, but they are documented in the [Analysts' Guide](https://dfe-analytical-services.github.io/analysts-guide/learning-development/r.html):

- **Cannot restore from an old `renv.lock`** when several older package versions fail to build. Assuming you're on the right version of R for the lockfile, and have ensured RTools is installed as described on this page then have a look at the analysts' guide, which covers using `renv::install()` and `renv::record()` to update individual packages, or `renv::init()` to discard the old lockfile entirely.
- **`shinytest2` errors about an invalid path to Chrome, or about old headless mode.** The guide explains setting `CHROMOTE_CHROME` and `CHROMOTE_headless` in your `.Renviron`.
- **SSL certificate errors from R or Python.** If a request is being blocked by the DfE firewall, contact the [Statistics Development Team](mailto:statistics.development@education.gov.uk) with the command and error so we can escalate it to the Network Operations team.

If you have run through everything here and are still seeing problems, please get in touch with the [Statistics Development Team](mailto:statistics.development@education.gov.uk) with the output of `diagnostic_test(full = TRUE)` attached.
